High-Dimensional Regularized Discriminant Analysis

### 2015-10-24 reports/paper-hdrda.Rnw

We consider regularized discriminant analysis (\emph{RDA}) from a classic paper
by \cite{Friedman:1989tm}, who proposed improvements to the \emph{QDA}
classifier by employing a covariance matrix estimator that partially pools the
individual sample covariance matrices with the pooled sample covariance matrix
from the \emph{LDA} classifier. Given that the assumption of equal population
covariance matrices for the \emph{LDA} classifier is often too stringent in
practice \citep{Clemmensen:2011kr}, the \emph{RDA} classifier's partial pooling
can relax the linearity assumption by instead incorporating unequal covariance
matrices while retaining as special cases the \emph{LDA} and QDA
classifiers. Covariance-matrix regularization is then applied to the partially
pooled covariance-matrix estimator by shrinking it towards a scaled identity
matrix to ensure positive definiteness.

We reparameterize the \emph{RDA} classifier similar to \cite{Hastie:2008dt} and
\cite{Halbe:2007ea} and refer to our parameterization as the high-dimensional
\emph{RDA} (\emph{HDRDA}) classifier. First, we show the pooling parameter in
the \emph{HDRDA} classifier determines the contribution of each training
observation to the estimation of each class covariance matrix, enabling
interpretability that has been previously lacking with the \emph{RDA} classifier
\citep{Bensmail:1996fw}. Next, applying a reasoning similar to \cite{Ye:2006tq},
we establish that the matrix operations corresponding to the null space of the
pooled sample covariance matrix are redundant and can be discarded from the
\emph{HDRDA} decision rule without loss of classificatory information. Moreover,
because the \emph{HDRDA} decision rule is invariant to adjustments to the zero
eigenvalues, the decision rule in the original feature space is equivalent to a
decision rule in a much lower dimension (i.e., the intrinsic dimension), such
that the matrix inverses and determinants of relatively small matrices can be
rapidly computed. Thus, the dimension reduction inherent in the \emph{HDRDA}
classifier yields a substantial reduction in computation during model
selection. Also, as a result, we show that a large number of shrinkage methods
that are special cases of the \emph{HDRDA} classifier have no effect on the zero
eigenvalues of the covariance-matrix estimators. Such methods include work from
\cite{Srivastava:2007ww}, \cite{Rao:1971ul}, and other methods studied by
\cite{Ramey:2013ji} and \citet*{Xu:2009fl}.

We provide an efficient algorithm that expedites the model-selection process for
the \emph{HDRDA} classifier and then compare its computational runtime with an
implementation of the \emph{RDA} classifier from the {\tt klaR} R package to
illustrate that the \emph{HDRDA} classifier greatly reduces computation when $p
\gg N$. We then study the classification performance of the \emph{HDRDA}
classifier on three artificial and four real high-dimensional data sets. We
demonstrate that the \emph{HDRDA} classifier often attains superior
classification accuracy to several recent classifiers designed for small-sample,
high-dimensional data from \cite{Tong:2012hw}, \cite{Witten:2011kc},
\cite{Pang:2009ik}, and \cite{Guo:2007te}. We also include as a benchmark the
random forest from \cite{Breiman:2001fb} because \cite{FernandezDelgado:2014ul}
has concluded that the random forest is often superior to other classifiers in
benchmark studies.

@@

A primary contributing factor to the efficiency of Algorithm
\ref{alg:hdrda_model_selection} is our usage of the compact singular value
decomposition (SVD) (sometimes called the reduced or fast SVD). Rather than
computing the eigenvalue decomposition of $\widehat{\bm \Sigma}$ to obtain $\bm
U_1$, we instead obtain $\bm U_1$ by computing the eigendecomposition of a much
smaller $N \times N$ matrix when $p \gg N$
\cite[Chapter~18.3.5]{Hastie:2008dt}. Applying the SVD, we decompose $\bm X_c =
\bm M \bm \Delta \bm U^{T}$, where $\bm M \in \mathbb{R}_{N \times p}$ is
orthogonal, $\bm \Delta \in \mathbb{R}_{p \times p}^{\ge}$ is a diagonal matrix
consisting of the singular values of $\bm X_c$, and $\bm U \in \mathbb{R}_{p
  \times p}$ is orthogonal. Recalling that $\widehat{\bm \Sigma} = N^{-1} \bm
X_c^{T} \bm X_c$, we have the eigendecomposition $\widehat{\bm \Sigma} = \bm U
\bm D \bm U^{T}$, where $\bm U$ is the matrix of eigenvectors of $\widehat{\bm
  \Sigma}$ and $\bm D = N^{-1} \bm \Delta$ is the diagonal matrix of eigenvalues
of $\widehat{\bm \Sigma}$. Now, we can obtain $\bm M$ and $\bm D$ efficiently
from the eigenvalue decomposition of the $N \times N$ matrix $\bm X_c \bm
X_c^{T} = \bm M \bm D \bm M^{T}$. Next, we compute $\bm U = \bm X_c^{T} \bm M
\bm D^{+/2}$, where $\bm D^{+/2} = \bm D_q^{-1/2} \oplus \bm 0_{N-q}$. We then
determine $q$, the number of numerically nonzero eigenvalues present in $\bm D$,
by calculating the number of eigenvalues that exceeds some tolerance value, say,
$\epsilon = 1 \times 10^{-6}$. After calculating $q$, we extract $\bm U_1$ by
retaining only the first $q$ columns of $\bm U$.

@@

In this section, we compare our proposed classifier with four classifiers
recently proposed for small-sample, high-dimensional data along with the
random-forest classifier from \cite{Breiman:2001fb} as a benchmark. Our
competitors include shrunken centroids regularized discriminant analysis from
\cite{Guo:2007te} and penalized linear discriminant analysis from
\cite{Witten:2011kc}. We also consider two modifications of diagonal linear
discriminant analysis from \cite{Tong:2012hw} and \cite{Pang:2009ik}, where the
former employs an improved mean estimator and the latter utilizes an improved
variance estimator. Below, we refer to each classifier by the first author's
surname.\footnote{TODO: Consider adding a heatmap of error rates of a single
  $(\lambda, \gamma)$ grid.}

We used version 3.2.2 of the open-source statistical software {\tt R} for the
classification study. We used the implementations of the Witten and Guo
classifiers from the {\tt penalizedLDA} and {\tt rda} packages, respectively.
Because the {\tt rda} package does not perform the authors' ``Min-Min'' rule
automatically, we applied this rule within our {\tt R} code. We also utilized
the implementations of the Pang and Tong classifiers available in the {\tt
  sparsediscrim} package. We used the implementation of the random-forest
classifier from the {\tt randomForest} package. For each classifer we explicitly
set prior probabilities as equal if applicable. All other classifier options
were set to their default settings.

>>>>

Here, we present a new classification method intended for the $p > N$ case that
builds on the classic regularized discriminant analysis (\emph{RDA}) proposed by
\cite{Friedman:1989tm}. Our proposed high-dimensional \emph{RDA} (\emph{HDRDA})
classifier employs a biased covariance matrix estimator that partially pools the
individual sample covariance matrices from the \emph{QDA} classifier with the
pooled sample covariance matrix from the \emph{LDA} classifier. We then shrink
the resulting covariance-matrix estimator towards a scaled identity matrix to
ensure positive definiteness. By reparameterizing the \emph{RDA} classifier
similar to \cite{Hastie:2008dt} and \cite{Halbe:2007ea}, we first show the
pooling parameter in the \emph{HDRDA} classifier determines the contribution of
each training observation to the estimation of each class covariance matrix,
enabling interpretability that has been previously lacking with the \emph{RDA}
classifier \citep{Bensmail:1996fw}.

Next, we achieve a substantial reduction in dimension to reduce greatly the
computation during model selection by applying a reasoning similar to
\cite{Ye:2006tq}. We establish that the matrix operations corresponding to the
null space of the pooled sample covariance matrix are redundant and can be
discarded from the \emph{HDRDA} decision rule without loss of classificatory
information. Moreover, because the \emph{HDRDA} decision rule is invariant to
adjustments to the zero eigenvalues, the decision rule in the original feature
space is equivalent to a decision rule in a much lower dimension, such that the
matrix inverses and determinants of relatively small matrices can be rapidly
computed. Furthermore, we show that a large number of shrinkage methods that are
special cases of the \emph{HDRDA} classifier have no effect on the zero
eigenvalues of the covariance-matrix estimators. Such methods include work from
\cite{Srivastava:2007ww}, \cite{Rao:1971ul}, and other methods studied by
\cite{Ramey:2013ji} and \citet*{Xu:2009fl}.

We provide an efficient algorithm that expedites the model-selection process for
the \emph{HDRDA} classifier when $p \gg N$. We then study the classification
performance of the \emph{HDRDA} classifier on four real high-dimensional data
sets along with a simulation experiment that generalizes that initially
conducted by \cite{Guo:2007te}. We demonstrate that the \emph{HDRDA} classifier
often attains superior classification accuracy to several recent classifiers
designed for small-sample, high-dimensional data from \cite{Tong:2012hw},
\cite{Witten:2011kc}, \cite{Pang:2009ik}, and \cite{Guo:2007te}. We also include
as a benchmark the random forest from \cite{Breiman:2001fb} because
\cite{FernandezDelgado:2014ul} has concluded that the random forest is often
superior to other classifiers in benchmark studies.

@@

A primary contributing factor to the efficiency of Algorithm
\ref{alg:hdrda_model_selection} is our usage of the compact singular value
decomposition (SVD) (sometimes called the reduced or fast SVD). Rather than
computing the eigenvalue decomposition of $\widehat{\bm \Sigma}$ to obtain $\bm
U_1$, we instead obtain $\bm U_1$ by computing the eigendecomposition of a much
smaller $N \times N$ matrix when $p \gg N$
\cite[Chapter~18.3.5]{Hastie:2008dt}. Applying the SVD, we decompose $\bm X_c =
\bm M \bm \Delta \bm U^{T}$, where $\bm M \in \mathbb{R}_{N \times p}$ is
orthogonal, $\bm \Delta \in \mathbb{R}_{p \times p}^{\ge}$ is a diagonal matrix
consisting of the singular values of $\bm X_c$, and $\bm U \in \mathbb{R}_{p
  \times p}$ is orthogonal. Recalling that $\widehat{\bm \Sigma} = N^{-1} \bm
X_c^{T} \bm X_c$, we have the eigendecomposition $\widehat{\bm \Sigma} = \bm U
\bm D \bm U^{T}$, where $\bm U$ is the matrix of eigenvectors of $\widehat{\bm
  \Sigma}$ and $\bm D = N^{-1} \bm \Delta$ is the diagonal matrix of eigenvalues
of $\widehat{\bm \Sigma}$. Now, we can obtain $\bm M$ and $\bm D$ efficiently
from the eigenvalue decomposition of the $N \times N$ matrix $\bm X_c \bm
X_c^{T} = \bm M \bm D \bm M^{T}$. Next, we compute $\bm U = \bm X_c^{T} \bm M
\bm D^{+/2}$, where $\bm D^{+/2} = \bm D_q^{-1/2} \oplus \bm 0_{N-q}$. We then
determine $q$, the number of numerically nonzero eigenvalues present in $\bm D$,
by calculating the number of eigenvalues that exceeds some tolerance value, say,
$1 \times 10^{-6}$. After calculating $q$, we extract $\bm U_1$ by retaining
only the first $q$ columns of $\bm U$.

@@

% TODO: Combine first two paragraphs -- list R package along with authors

In this section, we compare our proposed classifier with four classifiers
recently proposed for small-sample, high-dimensional data. Our
competitors include shrunken centroids regularized discriminant analysis from
\cite{Guo:2007te} and penalized linear discriminant analysis from
\cite{Witten:2011kc}.

We used the implementations of the Witten and Guo
classifiers from the {\tt penalizedLDA} and {\tt rda} packages, respectively.
Because the {\tt rda} package does not perform the authors' ``Min-Min'' rule
automatically, we applied this rule within our {\tt R} code.

We also consider two modifications of diagonal linear discriminant analysis from
\cite{Tong:2012hw} and \cite{Pang:2009ik}, where the former employs an improved
mean estimator and the latter utilizes an improved variance estimator. We also
utilized the implementations of the Pang and Tong classifiers available in the
{\tt sparsediscrim} package.

We also included the random forest from \cite{Breiman:2001fb} as a benchmark
based on the findings of \cite{FernandezDelgado:2014ul}, who has concluded that
the random forest is often superior to other classifiers in benchmark studies.
We used the implementation of the random-forest classifier from the {\tt
  randomForest} package with 250 trees and 100 maximum nodes.

For each classifer we explicitly set prior probabilities as equal if
applicable. All other classifier options were set to their default settings.
We used version 3.2.2 of the open-source statistical software {\tt R} for the
classification study.  Below, we refer to each classifier by the first author's
surname.
